\subsection{Algorithms parameterized by neighborhood diversity}

\begin{definition}[Neighborhood diversity]
    A graph $G = (V,E)$ has neighborhood diversity $\nd(G) = d$ if it can be $d$-partitioned in $\{D_1, \dots, D_d\}$, with each $D_i$ being a type class of $G$.
\end{definition}

As previously discussed, a type is a maximal set of vertices that are either true or false twins to each other.
If $D_i$ is composed of true twins, we say that it is a \tdef{true twin class} $T_i$ and, clearly, $G[D_i]$ is clique.
Similarly, if $D_i$ is composed of false twins, it is a \tdef{false twin class} $F_i$ and $G[D_i]$ is an independent set.
When $|D_i| = 1$, we treat the class differently depending on the problem.

\subsubsection{Biclique coloring}

For \textsc{biclique coloring}, if there is some $D_i$ with a single vertex, we shall treat it as a true twin class.

\begin{observation}
    \label{obs:biclique_true_twins}
    Given $G$ and a true twin class $T_i$ of $G$, any $k$-biclique-coloring $\varphi$ of $G$ has $|\varphi(T_i)| = |T_i|$.
\end{observation}

\begin{lemma}
    \label{lem:biclique_false_twins}
    Given $G$ and a false twin class $F \subset V(G)$, any  $k$-biclique-coloring $\varphi'$ of $G$ can be changed into a $k$-biclique-coloring $\varphi$ of $G$ such that $|\varphi(F)| \leq 2$.
\end{lemma}

\begin{tproof}
    If $|\varphi'(F)| \leq 2$, $\varphi = \varphi'$.
    Otherwise, there exists $f_1, f_2, f_3 \in F$ with three different colors. 
    Since every maximal biclique $B$ of $G$ with at least one element of $F$ has that $F \subset B$ and thus $|\varphi'(B)| \geq 3$.
    By making $\varphi(f_1) = \varphi'(f_1)$ and $\varphi(f_3) = \varphi(f_2) = \varphi'(f_2)$, we obtain $|\varphi(B)| \geq |\varphi(F)| \geq 2$.
    Repeating this process until $|\varphi(F)| = 2$ does not make any biclique monochromatic and completes the proof.
\end{tproof}

The central idea of our parameterized algorithm is to build an induced subgraph $H$ of $G$ and, afterward, use the results established here and in Section~\ref{sec:biclique_exact} to show that the solution to a particular instance of \textsc{set multicover} derived from $H$ can be transformed in a solution to \textsc{biclique coloring} of $G$.


\begin{definition}[B-Projection and B-Lifting]
    Let $T_i$ and $F_j$ be as previously discussed.
    We define the following projection rules:
    $\forall t_i^q \in T_i,\ \Proj{B}(t_i^q) = \{t'_i\}$;
    for $f_j^1\in F_j,\ \Proj{B}(f_j^1) = \{f'{_j^1}\}$;
    $\forall f_j^r \in F_j \setminus \{f_j^1\}$, $\Proj{B}(f_j^r) = \{f'{_j^2}\}$
    and $\Proj{B}(X) = \bigcup_{u \in X} \Proj{B}(u)$.
    
    Lifting rules are defined as $\Lift{B}(t'_i) = \{t_i\}$; $\Lift{B}(f'{_j^1}) = \{f_j^1\}$; $\Lift{B}(f'{_j^2}) = F_j \setminus \{f_j^1\}$ and $\Lift{B}(Y) = \bigcup_{u \in Y} \Lift{B}(u)$. Note that $\Proj{B}(\Lift{B}(X)) = X$, for any $X$.
\end{definition}


\begin{definition}[B-Projected Graph]
    The B-projected graph $H$ of $G$ is defined as $V(H) = \Proj{B}(V(G))$ and $v'_iv'_j \in E(H)$ if and only if there exist $v_i \in \Lift{B}(v'_i)$ and $v_j \in \Lift{B}(v'_j)$ such that $v_iv_j \in E(G)$. $H$ is an induced subgraph of $G$.
\end{definition}

\begin{figure}[!htb]
    \centering
        \begin{tikzpicture}[scale=1]
            \GraphInit[unit=3,vstyle=Normal]
            \SetVertexNormal[Shape=circle, FillColor=black, MinSize=2pt]
            \tikzset{VertexStyle/.append style = {inner sep = \inners, outer sep = \outers}}
            \SetVertexNoLabel
            \Vertex[x=0,y=0]{a}
            \Vertex[x=-1,y=0]{b}
            \Vertex[x=0,y=-1]{c}
            
            
            
            \Vertex[x=1,y=0]{h}
            \Vertex[x=0,y=1]{i}
            \Vertex[a=45, d=1]{j}
            
            \Edge(a)(b)
            \Edge(a)(c)
            
            \Edge(a)(h)
            \Edge(a)(i)
            \Edge(a)(j)
            \begin{scope}[shift={(-1.65cm, 0cm)}]
                \grComplete[RA=0.65]{3}
            \end{scope}
            \begin{scope}[shift={(0cm, -1.65cm)}]
                \grComplete[RA=0.65]{4}
            \end{scope}
        \end{tikzpicture}
    \hfill
        \begin{tikzpicture}[scale=1]
            \GraphInit[unit=3,vstyle=Normal]
            \SetVertexNormal[Shape=circle, FillColor=black, MinSize=2pt]
            \tikzset{VertexStyle/.append style = {inner sep = \inners, outer sep = \outers}}
            \SetVertexNoLabel
            \Vertex[x=0,y=0]{a}
            \Vertex[x=-1,y=0]{b}
            \Vertex[x=0,y=-1]{c}
            \Vertex[x=0,y=-1.65]{d}
            \Vertex[x=-1.65,y=0]{e}
            
            
            
            \Vertex[x=1,y=0]{h}
            \Vertex[x=0,y=1]{i}
            
            \Edge(a)(b)
            \Edge(a)(c)
            
            \Edge(a)(h)
            \Edge(a)(i)
            \Edge(b)(e)
            \Edge(c)(d)
        \end{tikzpicture}
    \hfill
    
    
    \caption{A graph (left), and its B-projected graph (right).}
    \label{fig:b-projected}
\end{figure}


For the remainder of this section, $G$ will be the input graph to \textsc{biclique coloring} and $H$ the B-Projected graph of $G$.
Our \textsc{set multicover} instance consists of $V(H)$ as the ground set, $\transs{B}(H)$ as the covering family, the size $k$ of the cover the same as the coloring of $G$ and $c(t_i') = |T_i|$ for every true twin class $T_i$ and $c(f'{_j^1}) = c(f'{_j^2}) = 1$ for each false twin class $F_j$.

The next observation follows directly from the fact that $\transs{B}(H)$ is closed under the subset operation, while the subsequent results allow us to move freely between $\transs{B}(G)$ and $\transs{B}(H)$.

\begin{observation}
    \label{obs:fatless_multicover}
    If there is a minimum $k$-multicover $\psi$ of $V(H)$ by $\transs{B}(H)$, then there exists a minimum $k$-multicover $\psi' = \{\psi_1, \dots, \psi_k\}$ such that $\left|\left\{j \mid u \in \psi_j\right\}\right| = c(u)$ for every $u \in V(H)$.
\end{observation}

\begin{lemma}
    \label{lem:lift_proj_biclique}
    If $B' \in \biq(H)$ then $B = \Lift{B}(B') \in \biq(G)$. Conversely, if $B \in \biq(G)$ and $B$ is not contained in any true twin class, then $B' = \Proj{B}(B) \in \biq(H)$.
\end{lemma}

\begin{tproof}
    ($\Rightarrow$) Note that $B$ is a biclique by the definition of $\Lift{B}$ and the fact that $B'$ is a biclique. By the contrapositive, suppose that $B \notin \biq(G)$ and that $u \in V(G)$ is such that $B \cup \{u\}$ is a (not necessarily maximal) biclique of $G$. Note that either:
    (i) if $u \in F_j$ then $F_j \nsubseteq B$ and $\Proj{B}(u) \notin B'$, because $u \notin \Lift{B}(f'{_j^1})$ or $u \notin \Lift{B}(f'{_j^2})$;
    or (ii) if $u \in T_i$ then $T_i \cap B = \emptyset$, which implies that $\Proj{B}(u) \notin B'$.
    %Since $H$ is an induced subgraph of $G$, no new bicliques can be created in $H$ that were not present in $G$.
    Since $B \cup \{u\}$ is a biclique, $\Proj{B}(u)$ is adjacent to only one partition of $B'$.
    The fact that $\Proj{B}(u) \notin B'$ implies that $\Proj{B}(B \cup \{u\}) = \Proj{B}(B) \cup \Proj{B}(u) = B' \cup \Proj{B}(u)$ is a biclique of $H$ and $B'$ is not maximal.
    
    ($\Leftarrow$) By the definition of $\Proj{B}$, $B' = (X, Y)$ must be a biclique of $H$. By the contrapositive, there is $u' \in V(H)$ such that $B' \cup \{u'\}$ is a (not necessarily maximal) biclique of $H$, and let $u \in \Lift{B}(u')$. By the definition of $\Lift{B}$, it follows that $u$ can only be adjacent to one of partition of $B$, say $\Lift{B}(X)$. Therefore, $u' \in Y$ and, for each $v \in \Lift{B}(Y)$, $uv \notin E(G)$, otherwise there would be $v' \in \Proj{B}(v)$ with $u'v' \in E(H)$. Therefore, $B \cup \{u\}$ is a biclique of $G$ and $B$ is not maximal.
\end{tproof}


\begin{theorem}
    \label{thm:lifted_transversal}
    $X \subseteq V(H)$ is in $\transs{B}(H)$ if and only if $\Lift{B}(X) \in \transs{B}(G)$.
\end{theorem}

\begin{tproof}
    ($\Rightarrow$) Recall that $X \in \transs{B}(H)$ if and only if no maximal biclique of $H$ is contained in $X$.
    It is clear that, for every $B' \in \biq(H)$, $B' \nsubseteq X$ implies that $\Lift{B}(B') \nsubseteq \Lift{B}(X)$, since no two vertices of $H$ are lifted to the same vertex of $G$, and $\Lift{B}(B') \in \biq(G)$ due to Lemma~\ref{lem:lift_proj_biclique}.
    Moreover, no biclique of $G$ entirely contained in a true twin class can be a subset of $\Lift{B}(X)$. As such, $\Lift{B}(X)$ contains a maximal biclique $B$ only if $\Proj{B}(B) \subseteq X$ and $\Proj{B}(B) \notin \biq(H)$, which is impossible due to Lemma~\ref{lem:lift_proj_biclique} and the assumption that $B$ is maximal.
    
    ($\Leftarrow$) Taking the contrapositive, $X \notin \transs{B}(H)$ implies that there is some maximal biclique $B'$ of $H$ such that $B' \subseteq X$. This implies that $\Lift{B}(B') \subseteq \Lift{B}(X)$, and, since $\Lift{B}(B')$ is a maximal biclique of $G$ due to Lemma~\ref{lem:lift_proj_biclique}, it holds that $\Lift{B}(X)$ is not a complement of transversal of $G$.
\end{tproof}


\begin{theorem}
    \label{thm:lifted_multicover}
    $\psi$ is a $k$-multicover of $H$ if and only if $G$ is $k$-biclique-colorable.
\end{theorem}

\begin{tproof}
    Recall that a $k$-partition is a $k$-biclique-coloring if and only if all elements of the partition belong to $\transs{B}(G)$. By the construction of our set multicover instance, we have that, for each $\psi_i$, $\psi_i \in \transs{B}(H)$. By making $\varphi = \left\{\Lift{B}(\psi_1), \dots, \Lift{B}(\psi_k)\right\}$, and recalling Observation~\ref{obs:fatless_multicover}, we have that each vertex $u \in V(H)$ is covered exactly $c(u)$ times; moreover, since true twins appear multiple times and types are equivalence relations, we can attribute to each $t_i^q$ any of the $|T_i|$ colors available, as long as no two receive the same color.
    Therefore, $\varphi$, after properly allocating the true twin classes, is a $k$-partition of $V(G)$.
    Due to Theorem~\ref{thm:lifted_transversal}, every $\Lift{B}(\psi_i)$ is a complement of transversal and therefore $\varphi$ is a valid $k$-biclique-coloring of $G$.
    
    For the converse, we first make use of Lemma~\ref{lem:biclique_false_twins} to guarantee that every false twin class is in at most two color classes. In particular, if two colors are required we force $f_j^1$ to have the smallest color and $F_j \setminus \{f_j^1\}$ to have the other one.
    Afterwards, for every color class $\varphi_i$, we take $\psi_i = \Proj{B}(\varphi_i)$.
    Note that, each color class has at most one element of each $T_i$. Also, for each $F_j$ and any two distinct color classes $\varphi_l, \varphi_r$, $\Proj{B}(\varphi_l) \cap \Proj{B}(\varphi_r) \cap \Proj{B}(F_j) = \emptyset$, since $f_j^1$ has a different color from $F_j \setminus \{f_j^1\}$.
    These observations guarantee that $\Lift{B}(\psi_i) = \varphi_i$ and, because of Theorem~\ref{thm:lifted_transversal}, $\psi_i \in \transs{B}(H)$.
    Finally, $\psi = \{\psi_1, \dots, \psi_k\}$ will be a valid $k$-multicover of $H$ because every vertex of $V(H)$ will be covered the required amount of times.
\end{tproof}

Note that the size of the largest true twin class is exactly the largest coverage requirement $b$ of our \textsc{set multicover} instance. Moreover, since we need at least $b$ colors to biclique color $G$, it holds that $b \leq k$.

\begin{theorem}
    \label{thm:fpt_biclique}
    There exists an $\bigOs{(k+2)^{2\nd(G)}}$ time algorithm to verify if $G$ is $k$-biclique-colorable.
\end{theorem}

\begin{tproof}
        Start by computing the type partition of $G$ in $\bigO{n^3}$ time and building $H$ in $O(n + m)$.
        Afterwards, solve the corresponding \textsc{set multicover} instance in $\bigOs{(k+2)^{2d}}$ time using Theorem~\ref{thm:set_multicover} and lift the multicover using the construction described in the proof of Theorem~\ref{thm:lifted_multicover} in $\bigO{n}$.
\end{tproof}

\subsubsection{Clique coloring}

For \textsc{clique coloring}, a type class with a single vertex is treated as a false twin class.
Unlike \textsc{biclique coloring}, both true and false twin classes are well behaved, one of the reasons we get a much better algorithm for this problem.


\begin{lemma}
    \label{lem:clique_false_twins}
    Given $G$ and a false twin class $F \subset V(G)$, any  $k$-clique-coloring $\varphi'$ can be changed into a $k$-clique-coloring $\varphi$ such that $|\varphi(F)| = 1$.
\end{lemma}

    \begin{tproof}
    If $|\varphi'(F)| = 1$, we are done.
    Otherwise, there exists $f_1, f_2 \in F$ such that $\varphi'(f_1) \neq \varphi'(f_2)$.
    For every maximal clique $C_1$ where $f_1 \in C_1$, define $C' = C \setminus \{f_1\}$ and note that $C_2 = C' \cup \{f_2\}$ is also a maximal clique.
    Since $\varphi'$ is an  coloring $|\varphi'(C') \cup \{\varphi'(f_1)\}| \geq 2$.
    Therefore, making $\varphi(f_2) = \varphi(f_1) = \varphi'(f_1)$ does not make $|\varphi(C_2)| = 1$. 
    Repeating this until $|\varphi(F)| = 1$ does not make any clique that intercepts $F$ monochromatic and completes the proof.
    \end{tproof}
    
\begin{lemma}
    \label{lem:clique_true_twins}
    Given $G$ and  a true twin class $T \subseteq V(G)$, any  $k$-clique-coloring $\varphi'$ can be changed into a $k$-clique-coloring $\varphi$ such that $|\varphi(T)| \leq 2$.
\end{lemma}

\begin{tproof}
    If $|\varphi'(T)| \leq 2$, we are done.
    Otherwise, there exists $t_1, t_2, t_3 \in T$ with different colors.
    Note that, for every maximal clique $C$ that intercepts $T$, $C \subseteq T$.
    Therefore, $|\varphi'(C)| \geq |\varphi'(T)| \geq 3$.
    By making $\varphi(t_1) = \varphi'(t_1)$ and $\varphi(t_3) = \varphi(t_2) = \varphi'(t_2)$ we have $|\varphi(C)| \geq |\varphi(T)| \geq 2$.
    Repeating this process until $|\varphi(T)| \leq 2$ does not make any clique that intercepts $T$ monochromatic and completes the proof.
\end{tproof}

We use sightly different projection and lifting rules for this problem.
However, they still rely on the fact that a type has a constant bound on the number of colors.

\begin{definition}[C-Projection and C-Lifting]
    Let $T_i$ be any true twin class and $F_j$ be any false twin class.
    We define the following projection rules:
    for $t_i^1 \in T_i,\ \Proj{C}(t_i^1) = \{t'{_i^1}\}$,
    $\forall t_i^q \in T_i \setminus \{t_i^1\}$, $\Proj{C}(t_i^q) = \{t'{_i^2}\}$,
    $\forall f_j^r \in F_j$, $\Proj{C}(f_j^r) = \{f'_j\}$
    and $\Proj{C}(X) = \bigcup_{u \in X} \Proj{C}(u)$.
    
    Lifting rules are defined as $\Lift{C}(t'{_i^1}) = \{t_i^1\}$,
    $\Lift{C}(t'{_i^2}) = T_i \setminus \{t_i^1\}$,
    $\Lift{C}(f'_j) = \{f_j^1\}$ and $\Lift{C}(Y) = \bigcup_{u' \in Y} \Lift{C}(u')$. Note that $\Proj{C}(\Lift{C}(X)) = X$, for any $X$.
\end{definition}

\begin{definition}[C-Projected Graph]
    The C-projected graph $H$ of $G$ is defined as $V(H) = \Proj{C}(V(G))$ and $v'_iv'_j \in E(H)$ if and only if $\Lift{C}(v'_i)$ and $\Lift{C}(v'_j)$ are neighbors in $G$. Moreover, $H$ is an induced subgraph of $G$.
\end{definition}

\begin{figure}[!htb]
    \centering
        \begin{tikzpicture}[scale=1]
            \GraphInit[unit=3,vstyle=Normal]
            \SetVertexNormal[Shape=circle, FillColor=black, MinSize=2pt]
            \tikzset{VertexStyle/.append style = {inner sep = \inners, outer sep = \outers}}
            \SetVertexNoLabel
            \Vertex[x=0,y=0]{a}
            \Vertex[x=-1,y=0]{b}
            \Vertex[x=0,y=-1]{c}
            
            
            
            \Vertex[x=1,y=0]{h}
            \Vertex[x=0,y=1]{i}
            \Vertex[a=45, d=1]{j}
            
            \Edge(a)(b)
            \Edge(a)(c)
            
            \Edge(a)(h)
            \Edge(a)(i)
            \Edge(a)(j)
            \begin{scope}[shift={(-1.65cm, 0cm)}]
                \grComplete[RA=0.65]{3}
            \end{scope}
            \begin{scope}[shift={(0cm, -1.65cm)}]
                \grComplete[RA=0.65]{4}
            \end{scope}
        \end{tikzpicture}
    \hfill
        \begin{tikzpicture}[scale=1]
            \GraphInit[unit=3,vstyle=Normal]
            \SetVertexNormal[Shape=circle, FillColor=black, MinSize=2pt]
            \tikzset{VertexStyle/.append style = {inner sep = \inners, outer sep = \outers}}
            \SetVertexNoLabel
            \Vertex[x=0,y=0]{a}
            \Vertex[x=-1,y=0]{b}
            \Vertex[x=0,y=-1]{c}
            
            
            
            \Vertex[a=45, d=1]{j}
            
            \Edge(a)(b)
            \Edge(a)(c)
            
            \Edge(a)(j)
            \begin{scope}[shift={(-1.65cm, 0cm)}]
                \grComplete[RA=0.65]{3}
            \end{scope}
            \begin{scope}[rotate=90,shift={(-1.65cm, 0cm)}]
                \grComplete[RA=0.65]{3}
            \end{scope}
        \end{tikzpicture}
    \hfill
    
    
    \caption{A graph (left), and its C-projected graph (right).}
    \label{fig:c-projected}
\end{figure}


For the remainder of this section, $G$ will be the input graph to \textsc{clique coloring} and $H$ the C-Projected graph of $G$.


\begin{lemma}
    \label{lem:lift_proj_clique}
    If $C' \in \clq(H)$ then $\Lift{C}(C') \in \clq(G)$. Conversely, if $C \in \clq(G)$ then $\Proj{C}(C) \in \clq(H)$.
\end{lemma}

\begin{tproof}
    ($\Rightarrow$) Note that $C = \Lift{C}(C')$ is a clique due to the definition of $\Lift{C}$ and the fact that $C'$ is a clique.
    By the contrapositive, suppose that $C$ is not a maximal clique.
    In this case, there is some vertex $u \in V(G)$ such that $C \cup \{u\}$ is a (not necessarily maximal) clique of $G$. Note that either:
    (i) if $u \in T_i$, $T_i \nsubseteq C$ and $u \notin \Lift{C}(t'{_i^1})$ or $u \notin \Lift{C}(t'{_i^2})$, thus $\Proj{C}(u) \notin C'$;
    (ii) if $u \in F_j$, $F_j \cap C = \emptyset$ and $\Proj{C}(u) \notin C'$.
    Since no two vertices of $H$ are lifted to the same vertex of $G$ and $\Proj{C}(u) \notin C'$, it follows that  $\Proj{C}(C \cup \{u\}) = \Proj{C}(C) \cup \Proj{C}(u) = C' \cup \Proj{C}(u)$ is a clique by the definition of $\Proj{C}$.
    
    ($\Leftarrow$) Clearly, $C' = \Proj{C}(C)$ is a clique of $H$, due to the definition of $\Proj{C}$.
    Suppose, however, that $C' \notin \clq(H)$, which implies that there is some $u' \in V(H)$ such that $C' \cup \{u'\}$ is a clique of $H$ and let $u \in \Lift{C}(u')$.
    By the definition of $\Lift{C}$, $C \subseteq N(u)$ if and only if $\Proj{C}(C) \subseteq N(u')$, which implies that $C'$ is not maximal only if $C$ is not maximal.
    A contradiction that completes the proof.
\end{tproof}

\begin{theorem}
    \label{thm:projected_clique_coloring}
     $G$ is $k$-clique-colorable if and only if $H$ is $k$-clique-colorable.
\end{theorem}

\begin{tproof}
    Let $\varphi_G$ be a $k$-clique-coloring of $G$ that complies with Lemmas~\ref{lem:clique_false_twins} and~\ref{lem:clique_true_twins}.
    Without loss of generality, for every $T_i$, we color $t_i^1$ with one color and $T_i \setminus \{t_i^1\}$ with the other, if it exists, otherwise color every vertex of $T_i$ with the same color.
    We define the $k$-clique-coloring of $H$ as $\varphi_H(u') = \varphi_G(u \in \Lift{C}(u'))$, for every $u' \in V(H)$.
    Suppose now that there exists some $C' \in \clq(H)$ such that $|\varphi_H(C')| = 1$.
    By Lemma~\ref{lem:lift_proj_clique}, $\Lift{C}(C')$ is a maximal clique of $G$ and, since $|\varphi_G\left(\Lift{C}(C')\right)| = 1$, it holds that $\Lift{C}(C')$ is a monochromatic maximal clique of $G$ and $\varphi_G$ is not a valid $k$-clique-coloring, which contradicts the hypothesis.
        
    Now, let $\varphi_H$ be a $k$-clique-coloring of $H$, and define $\varphi_G(u) = \varphi_H(u' \in \Proj{C}(u))$.
    By assuming that there exists some $C \in \clq(G)$ such that $\left|\varphi_G(C)\right| = 1$ and using Lemma~\ref{lem:lift_proj_clique}, it is clear that $\left|\varphi_H(\Proj{C}(C))\right| = 1$ which is impossible, since $\varphi_H$ is a valid $k$-clique-coloring of $H$.
\end{tproof}

\begin{theorem}
    \label{thm:fpt_clique}
    There exists an $\bigOs{2^{2\nd(G)}}$ time algorithm for \textsc{clique coloring}
\end{theorem}

\begin{tproof}
    Start by computing the optimal type partition of $G$ in $\bigO{n^3}$ time and building $H$ in $\bigO{n + m}$.
    Afterwards, color $H$ in $\bigOs{2^{2d}}$ time using Theorem~\ref{thm:clique_color_algorithm} and lift the coloring using the construction described in the proof of Theorem~\ref{thm:projected_clique_coloring} in $\bigO{n}$.
\end{tproof}