\chapter{Finding Cuts of Bounded Degree}

Unlike coloring problems, cut problems enforce properties \textit{between} sets.
The most famous cut problem is, most likely, \pname{Minimum Cut} (or \pname{Min Cut}), where the objective is to find a subset of edges of minimum size whose removal disconnects a specified pair of vertices.
A celebrated theorem, known as the \textit{Max-Flow Min-Cut theorem}~\citep{ford_fulkerson}, a special case of strong linear programming duality~\citep{linear_optimization}, states that the capacity of a minimum $s,t$ is precisely the value of the maximum flow between $s$ and $t$; when all edges have unit capacity, we simply seek a cut of minimum cardinality.
Other cases of cut problems in the literature include, for instance, \pname{Maximum Cut}, \pname{Max Cut} for short.
As the name implies, the goal is to find an $s,t$ cut of the graph with the maximum number of edges between them; however, unlike \pname{Min Cut}, \pname{Max Cut} is $\NPH$, in fact, the value of the cannot be arbitrarily approximated by a polynomial time approximation scheme unless $\P = \NP$, i.e. it is $\APX$-$\Hard$.
Naturally, one can ask whether or not there is a set of edges whose removal causes the disconnection of some vertices, but \textit{not others}.
Some research has been dedicated to this topic under the name of \pname{Multicut-Uncut}, especially in terms of parameterized complexity~\citep{marx_treewidth_reduction,marx_separation}, with powerful meta-theorems guaranteeing fixed parameter tractability for any class whose membership is hereditary and decidable.
Aside from constraints on which vertices are separated or the size of the separation, some cutting problems impose restrictions on the ``shape" of the edges crossing the cut.
Such is the case of \pname{Matching Cut} and the generalizations we discuss further on.

After some additional definitions and related work, we open the discussion of this chapter with a brief digression on possible generalization directions for \pname{Matching Cut}; our main goal, however, is at generalizing several results for the \pname{$d$-Cut} problem, which, to the best of our knowledge, we are the firsts to formally describe and investigate.
First, by using a reduction inspired by Chvátal's~\citep{chvatal_matching_cut}, we show that for every $d \geq 1$, \pname{$d$-Cut} is \NP-hard even when restricted to $(2d+2)$-regular graphs and that, if $\Delta(G) \leq d+2$, finding a $d$-cut can be done in polynomial time. The degree bound in the \NP-hardness result is unlikely to be improved: if we had an \NP-hardness result for \pname{$d$-Cut} restricted to $(2d+1)$-regular graphs, this would disprove the conjecture about the existence of internal partitions on $r$-regular graphs~\citep{DeVos09,internal_partition_regular6,internal_partition_regular3_4} for $r$ odd, unless every problem in $\NP$ could be solved in \textit{constant} time.
Afterwards, we present a simple exact exponential algorithm that, for every $d \geq 1$, runs in time $\bigOs{c_d^n}$ for some constant $c_d < 2$, hence improving over the trivial brute-force algorithm running in time $\bigOs{2^n}$.

We then proceed to analyze the problem in terms of its parameterized complexity.
Section~\ref{sec:param} begins with a proof, using the treewidth reduction technique of \citep{marx_treewidth_reduction}, that \pname{$d$-Cut} is $\FPT$ parameterized by the maximum number of edges crossing the cut.
Afterwards, we present a dynamic programming algorithm for \pname{$d$-Cut} parameterized by treewidth running in time $\bigOs{2^{\tw(G)+1}(d+1)^{2\tw(G) + 2}}$; in particular, for $d=1$ this algorithm runs in time $\bigOs{8^{\tw(G)}}$ and improves the one given by \cite{matching_cut_structural} for \pname{Matching Cut}, running in time  $\bigOs{12^{\tw(G)}}$.
By employing the cross-composition framework of \cite{cross_composition}, and using a reduction similar to the one in~\citep{matching_cut_ipec}, we show that, unless $\NP \subseteq \coNP/\poly$, there is no polynomial kernel for \pname{$d$-Cut} parameterized simultaneously by the number of crossing edges, the maximum degree, and the treewidth of the input graph.
We then present a polynomial kernel and an $\FPT$ algorithm when parameterizing by the distance to cluster.
This polynomial kernel is our main technical contribution, and it is strongly inspired by the technique presented by \cite{matching_cut_ipec} for \pname{Matching Cut}. Finally, we give an $\FPT$ algorithm parameterized by the distance to co-cluster, denoted by $\dcc(G)$.
These results imply fixed-parameter tractability for \pname{$d$-Cut} parameterized by $\tau(G)$.

\input{capitulos/secoes_dcut/intro.tex}
\input{capitulos/secoes_dcut/generalizations.tex}
\input{capitulos/secoes_dcut/np.tex}
\input{capitulos/secoes_dcut/param.tex}
\input{capitulos/secoes_dcut/conclusion.tex}